Information-Theoretic Registration with Explicit Reorientation of Diffusion-Weighted Images

### 2021-01-07 Apps/Overleaf/JMIV DWI  Registration/dwiregistering.tex
a partial rewite around the math, not done yet.

\section{Introduction}
\francois{Attempt to start more "generically":}
In this work, we study registration problems for images of the type $I:\Omega\to \Ff(\PP^2)$ with $\Omega$ an open, bounded domain of $\RR^3$, 
the image domain, $\PP^2$ is the projective plane of directions in $\RR^3$ and $\Ff(\PP^2)$ is a suitable function space on $\PP^2$: each $I(\bsx)$ is a function  $I(\bsx):\PP^2\to\RR$ which measures a phenomenon along a three-dimensional direction. We may and will also reinterpret such an image $I$ as an image $\bI:\Omega\times\PP^2\to\RR$, $\bI(\bsx, \bsv) = I(\bsx)(\bsv)$. These functions can be used to represent orientation distribution functions often used in Diffusion Weighted Imaging (DWI).

@@

\section{Locally Orderless Directional Imaging}
\francois{LORD sounds cool:-) (maybe not in Dansih) and would allow us to defer DWI explanations at some other point.} 
\label{sec:lor_dwi}
\subsection{Notations}
We start by introducing a few necessary notations. $\Omega\subset \mathbb R^3$ is the
spatial domain of the images under consideration. A scalar image is a function
$\bm I:\Omega\to\mathbb R^3$. We assume that we can extend it on the whole $\mathbb R^3$,
for instance by extending it with 0. The projective space of directions of $\mathbb R^3$
is denoted by $\mathbb P^2$, and the unit sphere of $\mathbb R^3$ by $\mathbb S^2$.  We
will encounter \emph{spatio-directional} images
$\bm I:\Omega\times \mathbb P^2 \to \mathbb R$, which we similarly assume too be
extendable to $\mathbb R^3\times \mathbb P^2$. This is necessary in both cases in order to
define their spatial smoothing via convolution. We use the following elementary property:
As $\mathbb P^2$ can naturally be identified as the quotient $\mathbb S^2/\{\pm 1\}$ by
the antipodal symmetry, a function $f:\mathbb P^2\to \mathbb R$ can be lifted to an
antipodal symmetric function $\tilde{f}:\mathbb{S}^2\to \mathbb{R}$. Conversely, any
antipodal symmetric function $g:\mathbb S^2\to \mathbb R$ factors through $\mathbb P^2$.
A spatio-directional image can (and will) be lifted to an antipodal symmetric image
$\tilde{\bm I}:\Omega\times \mathbb S^2$:
$\tilde{\bm I}(\bm x, -\bm v) =\tilde{\bm I}(\bm x, \bm v)$.  Such a lift allows to represent $\PP^2$-functions as $\SS^2$ ones, 
with the benefit that $\SS^2$ is an oriented manifold, while $\PP^2$ is not. This means in particular that $\PP^2$ does not have a global volume form.
On the other hand, $\SS^2$ has, and we use the one induced by its standard Riemannian metric.

@@

\subsection{Recall on the LOR Framework}
The LOR framework defines the density estimates over three scales: The image scale
$\sigma$, the intensity scale $\beta$, and the integration scale $\alpha$. In the context
of scalar registration, for a transformation $\phi:\mathbb{R}^3\rightarrow\mathbb{R}^3$,
the estimated histogram $h$ and the corresponding density $p$ is computed as
\begin{align}
\label{eq:loih}
  h_{\beta\alpha\sigma}&(i,j|\phi,\bm x)=\\
                 &\int_{\mathbb R^3}P_\beta( \bm I_{\sigma}(\phi(\bm x))-i)
                   P_\beta( \bm J_{\sigma}(\bm x)-j)W_\alpha(\bm \tau-\bm x)d\bm \tau\nonumber \\
  \label{eq:loipdf}
  p_{\beta\alpha\sigma}&(i,j|\phi,\bm x)\simeq \frac{h_{\beta\alpha\sigma}(i,j|\phi,\bm x)}
                         {\int_{\Lambda^2} h_{\beta\alpha\sigma}(k,l|\phi,\bm x)dk\; dl}
\end{align} 
where $i,j\in[a_1, a_2] =: \Lambda$ are values in the image intensity range,
$\bm I_{\sigma}(\bm \phi(\bm x))=(\bm I*K_\sigma)(\bm \phi(\bm x))$ and
$ J_{\sigma}(\bm x)=(\bm J*K_\sigma)(\bm x)$ are images convolved with the kernel
$K_\sigma$ with standard deviation $\sigma$, $P_\beta$ is a Parzen-window of scale
$\beta$, and $W_\alpha$ is a Gaussian integration window of scale $\alpha$. The marginals
are trivially obtained by integration over the appropriate variable.  The LOR-approach to
similarity lets us use a set of generalized similarity measures, the linear and non-linear
\begin{align}\label{eq:sim}
  \mathcal{F}_{lin}=\int_{\Lambda^2}f(i,j)p(i,j)di\; dj\\ \mathcal{F}_{non-lin}=\int_{\Lambda^2}f(p(i,j))di\; dj
\end{align}
where the linear measure $f(i,j)$ includes e.g. sum of squared differences and Huber, and
the non-linear $f(p(i,j))$ includes e.g. MI, normalized MI (NMI), see
\cite{darknersporring2012pami} for details.

\subsection{The LORD framework}
\francois{LORD! I like it!}
This work addresses the estimation of the image similarity $\mathcal F$ of DWI in the
context of nonrigid registration an extension of our previous work
\cite{jensen2015locally}.
In this context, $\bI$ and $\bJ$ are spatio-directional signals. Specifically, DWI MR
attenuation signals at location $\bm x$, for a gradient direction $\bm v$, are modeled by
$S(\bm x,\bm v) = S_0(\bm x) e^{-b\bm I(\bm x,\bm v)}$ \ \cite{tao2006method} and apparent
diffusion coefficients volumes are given by
$\bm I(\bm x,\bm v)=-\frac{1}{b}\log\frac{S(\bm x,\bm v)}{S_0(\bm x)}$. Gradient
directions $\bm v$ belong to $\mathbb S^2$ but the diffusion is orientation-free such that
$\bm I(\bm x,\bm v)\approx \bm I(\bm x,-\bm v)$ which naturally defines a
spatio-directional image $\Omega\times \mathbb P^2 \to \mathbb R$.  In order to apply
LOR-DWI, the histogram and density estimates \Cref{eq:loih} and \cref{eq:loipdf} must be
extended to spatio-directional data, and the action of the spatial transformation on the
directions must be defined.

@@

\subsubsection{Density, orientation and transformation}
Because of spatial convolutions, we assume that $\phi$ can be extended to $\RR^3$, by assuming that it is the identity out
of a compact set $D$ containing $\Omega$.
In the following, we will often omit the parameter $\bsx$. We set $\Phi = (\phi,\psi)$,
and we write the joint histogram and density for similarity in image registration as
\begin{align}
  h_{\beta\alpha\sigma\kappa}&(i,j|\Phi,\bm x)=\\ 
                      &\int_{\mathbb R^3 \times S^2}P_\beta(\bm I_{\sigma\kappa}(\phi(\bm x),\psi(\bm v))-i)
                        P_\beta(\bm J_{\sigma\kappa}(\bm x,\bm v)-j)\nonumber \\
                      &W_\alpha(\bm \tau-\bm x)d\bm \tau\times d\bm v \nonumber \label{eq:densityreg} \\
                      &p_{\beta\alpha\sigma\kappa}(i,j|\Phi,\bm x)=\frac{h_{\beta\alpha\sigma\kappa}(i,j|\Phi,\bm x)}
                       {\int_{\Lambda^{2}} h_{\beta\alpha\sigma\kappa}(i,j| \Phi,\bm x)dl\;dk} \;\;.
\end{align}
Assume that $W$ is a Gaussian kernel with standard deviation $\alpha$. If we left
$\alpha\rightarrow\infty$, we would obtain full spatial integration and can ignore the
integration scale $\alpha$.

\subsection{LOR-DWI and Free Form Deformation}
In \cite{jensen2015locally}, we assumed that $\phi$ was a global affine transformation,
with $\psi_x = \psi$ being the projectivization of its linear part. In the present work,
we assume $\phi$ to be a nonrigid transformation, and we use Rueckert \textit{et al.\/}
framework \cite{rueckert1999nonrigid} where the transformation $\phi$ is given as a
hierarchical spline representation \emph{linearly} parameterized by a spatial grid of
control points ${\bm c}$. In the next subsection, we use its vectorized form, so that
$\phi = \bm B\bm c$, $\Phi_{\bm c} = (\phi_{\bm c},\psi_{\bm c})$. We seek the
transformation $\Phi^*_{\bm c}$ which maximizes the regularized NMI
\begin{align}
  \Phi_c^* = \operatorname*{arg\,max}_c \mathcal M(\Phi_c) = \operatorname*{arg\,max}_c \mathcal F(\bm I\circ\Phi_c,\bm J)+\mathcal S({\Phi_c})
    \label{eq:argmax}
\end{align}
The dependency of $\mathcal M$ in $\bm c$ is complex. The following dependency diagram
(Fig. \ref{fig:dependcygraph}) for the mutual information term illustrates it.
\begin{figure}[ht]

\begin{equation*}
\begin{tikzcd}[column sep=0.4cm,row sep=0.5cm]
  ~& NMI( \bm I_{\sigma\kappa}\circ\Phi_{\bm c},\bm J_{\sigma\kappa};\beta)&~\\
  H_{\bm I_{\sigma\kappa}\circ\Phi_{\bm c};\beta} \ar[ur]&  
  H_{\bm J_{\sigma\kappa};\beta}\ar[u] & 
  H_{\bm I_{\sigma\kappa}\circ\Phi_{\bm c}, \bm J_{\sigma\kappa};\beta}\ar[lu]\\
  p_{\beta,\bm I_{\sigma\kappa}\circ\Phi_{\bm c}}\ar[u]  & 
  p_{\beta,\bm J_{\sigma\kappa}}\ar[u]  & 
  p_{\beta,\bm I_{\sigma\kappa}\circ\Phi_{\bm c}, \bm J_{\sigma\kappa}}\ar[u]\\
  ~ & h_{\beta,\bm I_{\sigma\kappa}\circ\Phi_{\bm c},\bm J_{\sigma\kappa}}\ar[ul]\ar[u]\ar[ur]& ~\\
  \bm I_{\sigma\kappa}(\phi_{\bm c},\psi_{\bm c})\ar[ur]&\psi_{\bm c}\ar[l]&\bm J_{\sigma\kappa}\ar[ul]\\
  \boxed{\bm c}\ar[r ]& \phi_{\bm c}\ar[lu] \ar[u]&
  \ar[from=1-1,to=1-3,color=red,dash,line width=0.65pt,start
  anchor={[xshift=-2.2em,yshift=2ex]center},end
  anchor={[xshift=+3.1em,yshift=2ex]center}]
  % 
  \ar[from=1-1,to=4-1,color=red,dash,line width=0.65pt,start
  anchor={[xshift=-2.2em,yshift=2ex]center},end
  anchor={[xshift=-2.2em,yshift=-1.5ex]center}]
  %
  \ar[from=4-1,to=4-3,color=red,dash,line width=0.65pt,start
  anchor={[xshift=-2.2em,yshift=-1.5ex]center},end
  anchor={[xshift=+3.1em,yshift=-1.5ex]center}]
  %
  \ar[from=1-3,to=4-3,color=red,dash,line width=0.65pt,start
  anchor={[xshift=+3.1em,yshift=2ex]center},end
  anchor={[xshift=+3.1em,yshift=-1.5ex]center}]
  %
  \ar[from=5-1,to=5-2,color=blue,dash,line width=0.65pt,start
  anchor={[xshift=-2.6em,yshift=1.7ex]center},end
  anchor={[xshift=+0.9em,yshift=1.7ex]center}]
  %
  \ar[from=6-1,to=6-2,color=blue,dash,line width=0.65pt,start
  anchor={[xshift=-2.6em,yshift=-1.7ex]center},end
  anchor={[xshift=+0.9em,yshift=-1.7ex]center}]
  %
  \ar[from=5-1,to=6-1,color=blue,dash,line width=0.65pt,start
  anchor={[xshift=-2.6em,yshift=1.7ex]center},end
  anchor={[xshift=-2.6em,yshift=-1.7ex]center}]
  %
  \ar[from=5-2,to=6-2,color=blue,dash,line width=0.65pt,start
  anchor={[xshift=+0.9em,yshift=1.7ex]center},end
  anchor={[xshift=+0.9em,yshift=-1.7ex]center}]
\end{tikzcd}
\end{equation*}
\caption{Dependency graph of the nonrigid DWI registration between the moving image
  $\bm I$ and the target image $\bm J$, with normalized mutual information (NMI) as the
  similarity measure. The deformation is parameterized by $\bm c$ so that any change in
  $\bm c$ will eventually affect the total similarity between the two images. The red
  frame contains MI registration elements which are classical, while the blue frame
  contains the part which extends the LOR framework to the LOR-DWI.}
    \label{fig:dependcygraph}
\end{figure}

@@

We use quasi-Newton methods to compute an optimum of \Cref{eq:argmax}, in particular
L-BFGS\cite{} from \cite{schmidt2005minfunc}, and thus need to compute the gradient of the
similarity with respect to control points vector $\bm c$. Most of the required
calculations are provided in
\cite{rueckert1999nonrigid,darknersporring2012pami,jensen2015locally}, they correspond to
the red frame in Figure \ref{fig:dependcygraph}, a full formula for the gradient is long
and not informative for the present work. We therefore only describe how

@@

\subsection{Regularization}
% Francois: reformulated from Henrik's paragraph.
So far, we have not specified the form of the regularizer $\mathcal S(\Phi_{\bm c})$ in
\Cref{eq:argmax}.  The regularization has received little attention due to the inherent
regularization from the smooth kernels and the additional directional structure. However,
we found that the last steps in the hierarchical transformation model required some
regularization to keep the deformation stable, due to the high resolution of the
deformation field. We used a simple regularization that penalizes a non-uniform grid by
the squared difference between a point $\bm c_i$ and its direct neighbours. The control
points are organized as a family of $R$ grids, from coarse to fine resolution, and the
regularizer $\mathcal S(\Phi_{\bm c})$ is the sum
$\sum_{r = 1}^R \mathcal{S}^r(\Phi_{\bm c^r})$ at each resolution, where
$\bm c^r = (c^r_1,\dots c^r_{p_r})^T$ is the grid of control points at resolution level
$r$, and we denote by $N^r(i)$ the set of indices $j$ such that control point $c^r_j$ is
neighbor to control point $c^r_j$ and by $|N^r(i)|$ its cardinal. We set
\begin{equation}
    \label{eq:regularizer}
    \mathcal{S}^r(\Phi_{\bm c})= -\frac{\lambda_r}{2}\sum_{i=1}^{p_r} \|c^r_i - \frac{1}{|N^r(i)|}\sum_{j\in N^r(i)} c^r_j\|^2\;\; .
\end{equation}
$\lambda_r$ is a strictly positive parameter controlling the degree of smoothness. The negative sign in \Cref{eq:regularizer} comes from the fact that we \emph{maximize} the objective function. In order to compute the gradient of $\mathcal{S}^r(\Phi_{\bm c^r})$, we define series of linear mappings $T^r_i:\bm c\mapsto |N^r(i)|c^r_i - \sum_{j\in N^r(i)}c^r_j$. The regularizer in \Cref{eq:regularizer} can be rewritten as
\begin{equation}
\mathcal{S}^r(\Phi_{\bm c})=-\frac{\lambda_r}{2}\sum_{i=1}^{p_r}\frac{1}{|N^r(i)|^2}\|T^r_i\bm c^r\|^2
\end{equation}
and by classical manipulation, we obtain that
\begin{equation}
    \label{eq:reggradient}
    \nabla_{\bm c}\mathcal{S}^r(\Phi_{\bm c}) = -\lambda_r\sum_{i=1}^{p_r}\frac{1}{|N^r(i)|^2}T_i^{r*} T_i^r\bm c^r \;\; .
\end{equation}
$T^{r*}_i$ is the adjoint of $T^r_i$. Operator $-\sum_{i=1}^{p_r}\frac{1}{|N^r(i)|^2}T_i^{r*} T^r_i$ is a discrete Laplacian. Our sought gradient is
\begin{equation}
    \nabla_{\bm c}\mathcal{S}(\Phi_{\bm c}) = -\sum_{r = 1}^R\lambda_r\sum_{i=1}^{p_r}\frac{1}{|N^r(i)|^2} T_i^{r*} T^r_i\bm c^r.
\end{equation}
In this paper, we chose $\lambda_1=\dots=\lambda_R$.

>>>>

\section{Introduction}
\francois{Attempt to start more "generically":}
In this work, we study registration problems for images of the type $I:\Omega\to \Ff(\PP^2)$ with 
$\Omega$ an open, bounded domain of $\RR^3$, the image domain, $\PP^2$ is the projective plane of 
directions in $\RR^3$ and $\Ff(\PP^2)$ is a suitable function space on $\PP^2$: each $I(\bsx)$ is a 
function  $I(\bsx):\PP^2\to\RR$ which provides a measure of a  phenomenon along a three-dimensional 
direction. We may and will also reinterpret such an image $I$ as an image 
$\bI:\Omega\times\PP^2\to\RR$, $\bI(\bsx, \bsv) = I(\bsx)(\bsv)$. 
These functions can be used to represent orientation distribution functions used in Diffusion Weighted Imaging (DWI).

@@

\section{Locally Orderless  Directional Imaging}
\francois{so that it ends up as LORD, sounds cool:-) (maybe not in Dansih) and would allow us to defer DWI explanations at some other point.} 
\label{sec:lor_dwi}
\subsection{Notations}
$\Omega\subset \mathbb R^3$ is the
spatial domain of the images under consideration. A scalar image is a function
$\bm I:\Omega\to\mathbb R^3$. We assume that we can extend it on the whole $\mathbb R^3$,
for instance by extending it with 0. The projective space of directions of $\mathbb R^3$
is denoted by $\mathbb P^2$, and the unit sphere of $\mathbb R^3$ by $\mathbb S^2$.  We
will encounter \emph{spatio-directional} images
$\bm I:\Omega\times \mathbb P^2 \to \mathbb R$, which we similarly assume too be
extendable to $\mathbb R^3\times \mathbb P^2$. This is necessary in both cases in order to
define their spatial smoothing via convolution. We use the following elementary property:
As $\mathbb P^2$ is naturally identified as the quotient $\mathbb S^2/\{\pm 1\}$ by
the antipodal symmetry, a function $f:\mathbb P^2\to \mathbb R$ can be lifted to an
antipodal symmetric function $\tilde{f}:\mathbb{S}^2\to \mathbb{R}$. Conversely, any
antipodal symmetric function $g:\mathbb S^2\to \mathbb R$ factors through $\mathbb P^2$.
A spatio-directional image can and will be lifted to an antipodal symmetric image
$\tilde{\bm I}:\Omega\times \mathbb S^2$:
$\tilde{\bm I}(\bm x, -\bm v) =\tilde{\bm I}(\bm x, \bm v)$.  Such a lift allows to represent $\PP^2$-functions as $\SS^2$ ones, 
with the benefit that $\SS^2$ is an oriented manifold, while $\PP^2$ is not. This means in particular that $\PP^2$ does not have a global volume form.
On the other hand, $\SS^2$ has, and we use the one induced by its standard Riemannian metric.

@@

\subsection{Recall on the LOR Framework}
The LOR framework defines the density estimates over three scales: The image scale
$\sigma$, the intensity scale $\beta$, and the integration scale $\alpha$. In the context
of scalar registration, for a transformation $\phi:\mathbb{R}^3\rightarrow\mathbb{R}^3$,
the estimated histogram $h$ and the corresponding density $p$ is computed as
\begin{align}
\label{eq:loih}
  h_{\beta\alpha\sigma}&(i,j|\phi,\bm x)=\\
                 &\int_{\mathbb R^3}P_\beta( \bm I_{\sigma}(\phi(\bm x))-i)
                   P_\beta( \bm J_{\sigma}(\bm x)-j)W_\alpha(\bm \tau-\bm x)d\bm \tau\nonumber \\
  \label{eq:loipdf}
  p_{\beta\alpha\sigma}&(i,j|\phi,\bm x)\simeq \frac{h_{\beta\alpha\sigma}(i,j|\phi,\bm x)}
                         {\int_{\Lambda^2} h_{\beta\alpha\sigma}(k,l|\phi,\bm x)dk\; dl}
\end{align} 
where $i,j\in[a_1, a_2] =: \Lambda$ are values in the image intensity range,
$\bm I_{\sigma}(\bm \phi(\bm x))=(\bm I*K_\sigma)(\bm \phi(\bm x))$ and
$ J_{\sigma}(\bm x)=(\bm J*K_\sigma)(\bm x)$ are images convolved with the kernel
$K_\sigma$ with standard deviation $\sigma$, $P_\beta$ is a Parzen-window of scale
$\beta$, and $W_\alpha$ is a Gaussian integration window of scale $\alpha$. The marginals
are trivially obtained by integration over the appropriate variable.  The LOR-approach to
similarity lets us use a set of generalized similarity measures, the linear and non-linear
\begin{align}\label{eq:sim}
  \mathcal{F}_{lin}=\int_{\Lambda^2}f(i,j)p(i,j)di\; dj\\ \mathcal{F}_{non-lin}=\int_{\Lambda^2}f(p(i,j))di\; dj
\end{align}
where the linear measure $f(i,j)$ includes e.g. sum of squared differences and Huber, and
the non-linear\linebreak $f(p(i,j))$ includes e.g. MI, normalized MI (NMI), see
\cite{darknersporring2012pami} for details.

\subsection{The LORD framework}
\francois{LORD! of course some mismatch with the MICCAI paper...}
This work addresses the estimation of the image similarity $\mathcal F$ of DWI in the
context of nonrigid registration an extension of our previous work
\cite{jensen2015locally}.
In this context, $\bI$ and $\bJ$ are spatio-directional signals. Specifically, DWI MR
attenuation signals at location $\bm x$, for a gradient direction $\bm v$, are modeled by
$S(\bm x,\bm v) = S_0(\bm x) e^{-b\bm I(\bm x,\bm v)}$ \ \cite{tao2006method} and apparent
diffusion coefficients volumes are given by
$\bm I(\bm x,\bm v)=-\frac{1}{b}\log\frac{S(\bm x,\bm v)}{S_0(\bm x)}$. Gradient
directions $\bm v$ belong to $\mathbb S^2$ but the diffusion is orientation-free such that
$\bm I(\bm x,\bm v)\approx \bm I(\bm x,-\bm v)$ which naturally defines a
spatio-directional image $\Omega\times \mathbb P^2 \to \mathbb R$.  In order to apply
LOR-DWI, the histogram and density estimates \Cref{eq:loih} and \cref{eq:loipdf} must be
extended to spatio-directional data, and the action of the spatial transformation on the
directions must be defined.

@@

\subsubsection{Density, orientation and transformation}
Because of spatial convolutions, we assume that $\phi$ can be extended to $\RR^3$, by assuming that it is the identity out
of a compact set $D$ containing $\Omega$.
In the following, we will often omit the parameter $\bsx$. We set $\Phi = (\phi,\psi)$,
and we write the joint histogram and density for similarity in image registration as
\begin{multline}
h_{\bsI\circ\Phi,\bsJ;\sigma\kappa\beta\alpha}(i,j|\bsx)=\\ 
                      \int_{\mathbb R^3 \times \SS^2}P_\beta(\bm I_{\sigma\kappa}(\phi(\bm x),\psi_{\bsx}(\bm v))-i)
                        P_\beta(\bm J_{\sigma\kappa}(\bm x,\bm v)-j) \times\\
                      W_\alpha(\bm \tau-\bm x)d\bm \tau d\bm v \label{eq:densityreg}
\end{multline}
\vspace{-5mm}
\begin{multline}
 p_{\bsI\circ\Phi,\bsJ_{\sigma\kappa\beta\alpha};}(i,j|\bsx)=\frac{h_{\bsI\circ\Phi,\bsJ;\sigma\kappa\beta\alpha}(i,j|\bsx)}
                       {\int_{\Lambda^{2}} h_{\bsI\circ\Phi,\bsJ;\sigma\kappa\beta\alpha}(l,k|\bsx)dl\;dk}\;.
\end{multline}
If one assumes that $W_\alpha$ is a Gaussian kernel with standard deviation $\alpha$ and we leave
$\alpha\rightarrow\infty$, we obtain full spatial integration, and we can write the joint histogram
and joint probability density functions (PDF)
\begin{multline}
\label{eq:histo_fullspatial}
	h_{\bsI\circ\Phi,\bsJ;\sigma\kappa\beta}(i, j)=\\
	\int_{\RR^3\times\SS^2}\!\!\!\!\!P_\beta(\bsI_{\sigma\kappa}(\phi(\bsx),\psi_{\bsx}(\bsv))-i)
	P_\beta(\bsJ_{\sigma\kappa}(\bsx,\bsv)-j)\,d\bsx d\bsv
\end{multline}
\vspace{-5mm}
\begin{equation}
	p_{\bsI\circ\Phi,\bsJ,\sigma\kappa\beta}(i,j) = \frac{h_{\bsI\circ\Phi,\bsJ;\sigma\kappa\beta}(i, j)}
	{\int_{\Lambda^2}h_{\bsI\circ\Phi,\bsJ;\sigma\kappa\beta}(l,k)dl\;dk}\;.~~~~~~~\hfil
\end{equation}
Similar formulas are obtained single histograms $h_{\bsK;\sigma\kappa\beta}$ and corresponding 
PDF $p_{\bsK;\sigma\kappa\beta}$ with $K = \bsI\circ\Phi$ and $\bsJ$, 
they are just the appropriate marginals of the joint distribution.
This is the situation we consider in rest of this work.

\subsection{LORD and Free Form Deformations}
In \cite{jensen2015locally}, we assumed that $\phi$ was a global affine transformation,
with $\psi_{\bsx} = \psi$ being the projectivization of its linear part. In the present work,
we assume $\phi$ to be a nonrigid transformation, and we use Rueckert \textit{et al.\/}
framework \cite{rueckert1999nonrigid} where the transformation $\phi$ is given as a
hierarchical spline representation \emph{linearly} parameterized by a spatial grid of
control points ${\bm c}$. In the next subsection, we use its vectorized form, so that
$\phi = \bm B\bm c$, $\Phi_{\bm c} = (\phi_{\bm c},\psi_{\bm c})$. We seek the
transformation $\Phi^*_{\bsc}$ which maximizes the regularized NMI
\begin{align}
  \Phi_{\bsc^*} = \operatorname*{arg\,max}_{\bsc} \mathcal M(\Phi_{\bsc}) = \operatorname*{arg\,max}_{\bsc} \mathcal F(\bm I\circ\Phi_{\bsc},\bm J)+\mathcal S({\Phi_{\bsc}})
    \label{eq:argmax}
\end{align}

We use quasi-Newton methods to compute an optimum of \Cref{eq:argmax}, in particular
L-BFGS from \cite{schmidt2005minfunc}, and thus need to compute the gradient of the
similarity with respect to control points vector $\bm c$. 
The dependency of $\Mm$ in $\bsc$ appears deep inside its definition, 
this will be reflected in its differential w.r.t. $\bsc$. In the next paragraph, we provide some 
details of its computation for the similarity term, and in paragraph \ref{sec:reg}, 
we detail computations the regularisation term.

@@

The NMI term is defined as
	\text{NMI}(\bsI\circ\Phi_{\bsc}, \Jj) = \frac{H(\bsI\circ\Phi_{\bsc}) + H(\bsJ) }{H(\bsI\circ\Phi_{\bsc}, \bsJ)}.
\end{equation} where the different terms are differential entropies $H(p) = -\int p(t)\log p(t)\,dt$.
The following diagram (Fig. \ref{fig:dependcygraph}) exhibits the PDFs that are used in the differential 
entropies which form the NMI term and illustrates its dependency in $\bsc$, which occurs only through 
$\Phi_{\bsc}$. A full formula for the gradient is long and not informative for the present work. Actually, 
many of the required calculations are similar to the ones in  \cite{rueckert1999nonrigid,darknersporring2012pami,jensen2015locally}, 
they correspond to the red frame in Fig. \ref{fig:dependcygraph}, and are obtained by application of standard derivation rules.  
 The computations in this paper which differ from the aforementioned are the ones 
corresponding to the blue frame in Fig. \ref{fig:dependcygraph}, where the directional terms are "uncovered". 
\francois{This paragraph is IMHO poorly written...I should rewor(d/k) this part...}
\begin{figure}[ht]
\begin{equation*}
\begin{tikzcd}[column sep=0.4cm,row sep=0.5cm]
  ~& NMI( \bsI\circ\Phi_{\bsc},\bsJ;\sigma\kappa\beta)&~\\
  H_{\bm I_{\sigma\kappa}\circ\Phi_{\bm c};\beta} \ar[ur]&  
  H_{\bm J_{\sigma\kappa};\beta}\ar[u] & 
  H_{\bm I_{\sigma\kappa}\circ\Phi_{\bm c}, \bm J_{\sigma\kappa};\beta}\ar[lu]\\
  p_{\bm I_{\sigma\kappa}\circ\Phi_{\bm c};\beta}\ar[u]  & 
  p_{\bm J_{\sigma\kappa};\beta}\ar[u]  & 
  p_{\bm I_{\sigma\kappa}\circ\Phi_{\bm c}, \bm J_{\sigma\kappa};\beta}\ar[u]\\
  ~ & h_{\bm I_{\sigma\kappa}\circ\Phi_{\bm c},\bm J_{\sigma\kappa};\beta}\ar[ul]\ar[u]\ar[ur]& ~\\
  \bm I_{\sigma\kappa}(\phi_{\bm c},\psi_{\bm c})\ar[ur]&\psi_{\bm c}\ar[l]&\bm J_{\sigma\kappa}\ar[ul]\\
  \boxed{\bm c}\ar[r ]& \phi_{\bm c}\ar[lu] \ar[u]&
  \ar[from=1-1,to=1-3,color=red,dash,line width=0.65pt,start
  anchor={[xshift=-2.2em,yshift=2ex]center},end
  anchor={[xshift=+3.1em,yshift=2ex]center}]
  % 
  \ar[from=1-1,to=4-1,color=red,dash,line width=0.65pt,start
  anchor={[xshift=-2.2em,yshift=2ex]center},end
  anchor={[xshift=-2.2em,yshift=-1.5ex]center}]
  %
  \ar[from=4-1,to=4-3,color=red,dash,line width=0.65pt,start
  anchor={[xshift=-2.2em,yshift=-1.5ex]center},end
  anchor={[xshift=+3.1em,yshift=-1.5ex]center}]
  %
  \ar[from=1-3,to=4-3,color=red,dash,line width=0.65pt,start
  anchor={[xshift=+3.1em,yshift=2ex]center},end
  anchor={[xshift=+3.1em,yshift=-1.5ex]center}]
  %
  \ar[from=5-1,to=5-2,color=blue,dash,line width=0.65pt,start
  anchor={[xshift=-2.6em,yshift=1.7ex]center},end
  anchor={[xshift=+0.9em,yshift=1.7ex]center}]
  %
  \ar[from=6-1,to=6-2,color=blue,dash,line width=0.65pt,start
  anchor={[xshift=-2.6em,yshift=-1.7ex]center},end
  anchor={[xshift=+0.9em,yshift=-1.7ex]center}]
  %
  \ar[from=5-1,to=6-1,color=blue,dash,line width=0.65pt,start
  anchor={[xshift=-2.6em,yshift=1.7ex]center},end
  anchor={[xshift=-2.6em,yshift=-1.7ex]center}]
  %
  \ar[from=5-2,to=6-2,color=blue,dash,line width=0.65pt,start
  anchor={[xshift=+0.9em,yshift=1.7ex]center},end
  anchor={[xshift=+0.9em,yshift=-1.7ex]center}]
\end{tikzcd}
\end{equation*}
\caption{Dependency graph of the nonrigid DWI registration between the moving image
  $\bm I$ and the target image $\bm J$, with normalized mutual information (NMI) as the
  similarity measure. The deformation is parameterized by $\bm c$ so that any change in
  $\bm c$ will eventually affect the total similarity between the two images. The red
  frame contains MI registration elements which are classical, while the blue frame
  contains the part which extends the LOR framework to the LOR-DWI.}
    \label{fig:dependcygraph}
 We therefore only describe how

@@

\subsection{Regularization}
\label{sec:reg}
% Francois: reformulated from Henrik's paragraph.
So far, we have not specified the form of the regularizer $\mathcal S(\Phi_{\bm c})$ in
\Cref{eq:argmax}.  The regularization has received little attention due to the inherent
regularization from the smooth kernels and the additional directional structure. However,
we found that the last steps in the hierarchical transformation model required some
regularization to keep the deformation stable, due to the high resolution of the
deformation field. We used a simple regularization that penalizes a non-uniform grid by
the squared difference between a point $\bm c_i$ and its direct neighbours. The control
points are organized as a family of $R$ grids, from coarse to fine resolution, and the
regularizer $\mathcal S(\Phi_{\bm c})$ is the sum
$\sum_{r = 1}^R \mathcal{S}^r(\Phi_{\bm c^r})$ at each resolution, where
$\bm c^r = (c^r_1,\dots c^r_{p_r})^T$ is the grid of control points at resolution level
$r$, and we denote by $N^r(i)$ the set of indices $j$ such that control point $c^r_j$ is
neighbor to control point $c^r_j$ and by $|N^r(i)|$ its cardinal. We set
\begin{equation}
    \label{eq:regularizer}
    \mathcal{S}^r(\Phi_{\bm c})= -\frac{\lambda_r}{2}\sum_{i=1}^{p_r} \|c^r_i - \frac{1}{|N^r(i)|}\sum_{j\in N^r(i)} c^r_j\|^2\;\; .
\end{equation}
$\lambda_r$ is a strictly positive parameter controlling the degree of smoothness. The negative sign in \Cref{eq:regularizer} comes from the fact that we \emph{maximize} the objective function. In order to compute the gradient of $\mathcal{S}^r(\Phi_{\bm c^r})$, we define series of linear mappings $T^r_i:\bm c\mapsto |N^r(i)|c^r_i - \sum_{j\in N^r(i)}c^r_j$. The regularizer in \Cref{eq:regularizer} can be rewritten as
\begin{equation}
\mathcal{S}^r(\Phi_{\bm c})=-\frac{\lambda_r}{2}\sum_{i=1}^{p_r}\frac{1}{|N^r(i)|^2}\|T^r_i\bm c^r\|^2
\end{equation}
and by classical manipulation, we obtain that
\begin{equation}
    \label{eq:reggradient}
    \nabla_{\bm c}\mathcal{S}^r(\Phi_{\bm c}) = -\lambda_r\sum_{i=1}^{p_r}\frac{1}{|N^r(i)|^2}T_i^{r*} T_i^r\bm c^r \;\; .
\end{equation}
$T^{r*}_i$ is the adjoint of $T^r_i$. Operator $-\sum_{i=1}^{p_r}\frac{1}{|N^r(i)|^2}T_i^{r*} T^r_i$ is a discrete Laplacian. Our sought gradient is
\begin{equation}
    \nabla_{\bm c}\mathcal{S}(\Phi_{\bm c}) = -\sum_{r = 1}^R\lambda_r\sum_{i=1}^{p_r}\frac{1}{|N^r(i)|^2} T_i^{r*} T^r_i\bm c^r.
\end{equation}
In this paper, we chose $\lambda_1=\dots=\lambda_R$.
